---
title: "Book-campaign-performance-report"
author: "Benkly"
date: "2026-03-23"
output:
 html_document:
  toc: true
# output: pdf_document
# output: ioslides_presentation
# output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This report acts as a follow-on to the report titled `Book-sales-report`, which deals with a scenario where a data analyst has been tasked to provide a high-level financial performance report by a company which publishes textbooks related to learning the R programming language.

For this report, we imagine the company has provided additional sales information for the fiscal year of 2019. The company launched a campaign on July 1^st^ 2019 to promote their books and would like to know whether their is evidence to suggest the campaign was successful at increasing sales and improving review quality.

## Aims

* To summarize sales and review sentiment before and after the campaign start date to determine whether there is evidence to suggest the campaign was a success.

To do this, date/time manipulation and string manipulation methods are necessary.

## Dataset
The dataset was sourced from [Dataquest](https://app.dataquest.io/c/94/m/516/guided-project%3A-creating-an-efficient-data-analysis-workflow%2C-part-2/1/introduction?path).

### Data Dictionary
| Column                  | Description                                                                 |
|-------------------------|-----------------------------------------------------------------------------|
| date                    | Date of sale                                                                |
| user_submitted_review   | Text based review from the buyer                                            |
| title                   | The title of the purchased book                                             |
| total_purchased         | The quantity of books purchased                                             |
| customer_type           | Whether the customer is an Individual or Business                           |

### Loading Libraries and Importing the Data

```{r loading libraries and dataset}
# Loading libraries
library(conflicted)
library(tidyverse)
library(lubridate)

# Resolve filter conflict
conflicts_prefer(dplyr::filter)

# Reading the Data
sales_2019 <- read_csv("sales2019.csv")
```

## Exploring the Dataset

```{r understanding the data}

# Dimension of the Dataframe
dim(sales_2019)

# Column Names
cols <- colnames(sales_2019)
cols

# Global view of the data
glimpse(sales_2019)

# Missing Values
missing <- function(df) {
  missing_values <- rep(0, times=length(colnames(df)))
  names(missing_values) <- colnames(df)
  i <- 1
  
  for (col in colnames(df)) {
    missing_values[i] <- is.na(df[[col]]) %>% sum
    i <- i + 1
  }
  return(missing_values)
}

missing(sales_2019)
```

The dataset contains `r dim(sales_2019)[1]` records and `r dim(sales_2019)[2]` columns. There are `r missing(sales_2019)[2]` missing values in the `user_submitted_review` column and `r missing(sales_2019)[4]` missing values in the `total_purchased` column.

---

Since one of our primary concerns is with sales performance before and after the campaign date, we ideally do not want to discard records with missing data in the `total_purchased` field. We can instead impute the missing data with the mean as a best estimate for the missing values.

In the case of the missing values in the `user_submitted_review` field, we can discard any records with no review as we cannot impute this data.

## Handling Missing Values

``` {r handling missing data}
# Remove all records with no user review

sales_2019_clean <- sales_2019 %>% 
  filter(!is.na(user_submitted_review))

# Calculate imputation value

average_purchase_amount <- sales_2019_clean %>%
  filter(!is.na(total_purchased)) %>%
  pull(total_purchased) %>%
  mean

# Impute missing purchase amounts with the calculated value

sales_2019_clean <- sales_2019_clean %>%
  mutate(
    imputed_purchases = if_else(is.na(total_purchased), round(average_purchase_amount, 1),
                                total_purchased)
  )

# Check for missing values on the cleaned dataset
missing(sales_2019_clean)

# Dimensions of cleaned dataset
dim(sales_2019_clean)
```

From the original `r dim(sales_2019)[1]` records, 885 records (`r (885/dim(sales_2019)[1])*100`%) have been removed corresponding to missing user review fields, leaving `r dim(sales_2019_clean)[1]` records. The remaining 583 null values in the `total_purchased` field have been imputed using the average of the `total_purchased` field (with nulls removed). 

## Handling the Review Column

The `user_submitted_review` field is more complicated in that there is no specific format that the data must fit into. It is a free-form text field which users can input any sequence of ASCII characters. For this report, we would like to assign a sentiment to each review as either positive, negative or neutral.

```{r review handling}
unique_reviews <- sales_2019_clean %>%
  pull(user_submitted_review) %>%
  unique()

unique_reviews
```

It seems there are only a handful of unique reviews (this is a simplified case). Since there are a relatively small number of unique reviews, we can assign sentiment to each using a custom function.

```{r assigning sentiments to reviews}
sentiment <- function(review) {
  review <- str_to_lower(review)
  sentiment_class <- case_when(
    str_detect(review, "awesome") ~ "Positive",
    str_detect(review, "learned a lot") ~ "Positive",
    str_detect(review, "never") ~ "Positive",
    str_detect(review, "not recommend") ~ "Negative",
    str_detect(review, "not needed") ~ "Negative",
    str_detect(review, "hate") ~ "Negative",
    str_detect(review, "other books were better") ~ "Negative",
    TRUE ~ "Neutral" # captures reviews which do not contain any of the substrings listed above
  )
}

sales_2019_clean <- sales_2019_clean %>%
  mutate(
    sentiment = unlist(map(user_submitted_review, sentiment)),
    sentiment_score = case_when(
      sentiment == "Positive" ~ 1,
      sentiment == "Neutral" ~ 0,
      TRUE ~ -1
    )
  )

knitr::kable(head(sales_2019_clean, 20))
```

## Preparing the `date` Field

Before moving to segmenting the dataset by date, we first need to prepare the date data for use. As it stands currently, the date is expressed as a string representation. The **lubridate** package can be used to format and parse datetime data.

```{r date manipulation}
campaign_date <- "07-01-2019" %>% mdy %>% as.numeric() # unix-time representation of the campaign date

sales_2019_clean <- sales_2019_clean %>% 
  mutate(
  date = mdy(date),
  during_campaign = if_else(date >= campaign_date, TRUE, FALSE)
)

knitr::kable(head(sales_2019_clean))

# Save cleaned data
write.csv(sales_2019_clean, "sales2019clean.csv", row.names=FALSE)
```

We have successfully categorized each record as either pre- or post-campaign, assigned a sentiment (and sentiment score) to each record and have a measure of sales performance using the `imputed_purchases` field. The data is ready for analysis.

## Analysis

We can begin by aggregating the data pre- and post-campaign to get an insight into whether an improvement in sales and sentiment occurred.

```{r segmenting the data by pre/post campaign}

# create a mode function to find the statistical mode of the sentiment column
mode <- function(x) {
  unique_vals <- unique(x)
  unique_vals[which.max(tabulate(match(x, unique_vals)))]
}

campaign_performance_breakdown <- sales_2019_clean %>%
  group_by(during_campaign) %>%
  summarize(
    sentiment = mode(sentiment),
    avg_sentiment_score = round(mean(sentiment_score), 3),
    total_purchases = sum(imputed_purchases),
    avg_quantity_purchased_per_sale = round(mean(imputed_purchases), 2)
  )

knitr::kable(campaign_performance_breakdown)
```

* There is only a very marginal improvement in review sentiment during the period of time when the campaign was running. 

* The number of books purchased actually declined slightly during the campaign period. 

These changes are small, so will certainly be statistically insignificant. Taking the data as a whole, there is insufficient evidence to support that the campaign was successful at improving sentiment or purchase of books. 

It could be the case that some customers responded better to the campaign depending on whether they are an individual versus a business. We can sub-divide the data further to explore this:

```{r segmenting the data by pre/post campaign and customer type}

campaign_performance_breakdown_by_customer_type <- sales_2019_clean %>%
  group_by(during_campaign, customer_type) %>%
  summarize(.groups="drop_last",
    sentiment = mode(sentiment),
    avg_sentiment_score = round(mean(sentiment_score), 3),
    total_purchases = sum(imputed_purchases),
    avg_quantity_purchased_per_sale = round(mean(imputed_purchases), 2)
  ) %>%
  arrange(customer_type)

knitr::kable(campaign_performance_breakdown_by_customer_type)
```

* Businesses show a slight improvement in both sales and review sentiment comparing pre-campaign performance to post-campaign performance.

* There is a decline in review sentiment and sales for individual customers however.

---
Finally, we can breakdown the sentiment change and sales performance for specific titles:

```{r segmenting the data by pre/post campaign and title}

campaign_performance_breakdown_by_customer_type <- sales_2019_clean %>%
  group_by(during_campaign, title) %>%
  summarize(.groups="drop_last",
    sentiment = mode(sentiment),
    avg_sentiment_score = round(mean(sentiment_score), 3),
    total_purchases = sum(imputed_purchases),
    avg_quantity_purchased_per_sale = round(mean(imputed_purchases), 2)
  ) %>%
  arrange(title)

knitr::kable(campaign_performance_breakdown_by_customer_type)
```

We can draw some interesting conclusions from this breakdown:

* The general sentiment for `Fundamentals of R For Beginners` noticeably improved, however, the number of sales of the book declined.

* More copies of `R For Dummies` were sold post-campaign, though the review sentiments did not significantly change.

* Very few copies of `R Made Easy` sold during either period. A slight improvement in sentiment and sales is noted during the campaign, but due to the small sample size, should not be taken too seriously.

* `R vs Python: An Essay` performed considerably worse during the campaign period, both sentiment and sales declined.

* Sales of `Secrets Of R For Advanced Students` improved during the campaign, however, the sentiment of the reviews worsened.

* The book `Top 10 Mistakes R Beginners Make` is the second worst performing title sold by the company. We can note an improvement in review sentiment post-campaign, however, sales declined slightly.